---
title: "Earthquake Depth vs Magnitude — Covariance & Correlation in R"
author: "Faithwin"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A teaching exercise using R's built-in quakes dataset to explore the relationship between earthquake depth and magnitude using covariance and correlation.

## Dataset
The quakes dataset contains observations of 1,000 seismic events near Fiji recorded by the Harvard LASA seismograph network. Each row captures:

# Variable    Description
lat            Latitude of earthquake epicentre
long           Longitude of earthquake epicentre
depth          Depth of earthquake (km)
mag            Richter magnitude
stations       Number of stations that detected the event

```{r quakes-head, echo=FALSE}
data(quakes)
head(quakes)
#>      lat   long depth mag stations
#> 1 -20.42 181.62   562 4.8       41
#> 2 -20.62 181.03   650 4.2       15
#> 3 -26.00 184.10    42 5.4       43
```

## Visualisation

Using scatterplot 

```{r quakes-plot, echo=FALSE}
plot(quakes$depth, quakes$mag,
     xlab = "Depth (km)",
     ylab = "Magnitude",
     main = "Earthquake Depth vs Magnitude")
```
The plot shows a slight downward trend. This means the deeper earthquakes tend to have marginally lower surface magnitudes but with substantial scatter, hinting at a weak relationship.


We cannot decide using only the scatterplot. Therefore, we do a covariance check. 

```{r, echo=FALSE}
cov(quakes$depth, quakes$mag)
#> [1] -20.02209
```
The negative sign confirms that depth and magnitude move in opposite directions: as depth increases, magnitude tends to decrease. However, the raw value of -20.02 is difficult to interpret in isolation because covariance retains the original units (km × Richter scale), making cross-variable comparisons unreliable. So, we run correlation instead. 


```{r, echo=FALSE}
cor(quakes$depth, quakes$mag)
#> [1] -0.2306377
```
A value of −0.23 indicates a weak negative relationship. Depth explains only a small portion of the variation in magnitude — consistent with the scatter visible in the plot.Correlation standardises the covariance to a scale of −1 to +1, making it interpretable. 

## Full correlation Matrix

```{r quakes-matrix, echo=FALSE}
cor(quakes)
#>                lat        long       depth         mag    stations
#> lat      1.0000000  -0.3645440  0.0310258  -0.0504617  -0.0022206
#> long    -0.3645440   1.0000000  0.1444434  -0.1730673  -0.0535125
#> depth    0.0310258   0.1444434  1.0000000  -0.2306377  -0.0735151
#> mag     -0.0504617  -0.1730673 -0.2306377   1.0000000   0.8511824
#> stations -0.0022206 -0.0535125 -0.0735151   0.8511824   1.0000000
```
#Notable Findings 
mag ↔ stations = 0.85 — Strong positive correlation. More powerful earthquakes are detected by a greater number of stations, which makes physical sense: larger seismic events produce stronger signals that reach more sensors.

depth ↔ mag = −0.23 — Weak negative correlation. Deeper earthquakes tend to register slightly lower surface magnitudes.

lat ↔ long = −0.36 — Moderate negative correlation, reflecting geographic clustering of seismic activity in the Fiji–Tonga region.

depth ↔ stations = −0.07 — Near zero. Depth has almost no bearing on how many stations detect the event.


#Thank you. 